Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-10-28 paper/current/main.tex
rewrite math part and use innter product

Here, we study the skip-gram negative sampling, or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning models of language~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered}, networks~\autocite{grover2016node2vec, perozzi2014deepwalk, linzhuo2020hyperbolic} and trajectories\autocite{liu2016exploring, feng2017poi2vec, yao2018representing, cao2019habit2vec, crivellari2019motion, solomon2018predict}.
It works under the notion that a good representation should facilitate prediction, learning a mapping between words that can predict a target word based on its context (surrounding words).
The model is also computationally efficient, easy to use, robust to noise, and can encode relations between entities as geometric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis, hashimoto2016word}.
When applied to the trajectory data, each location is encoded into a single vector representation, and vectors relate to one another based on the likelihood of locations appearing adjacent to one another in the same trajectory. Also,  \textit{word2vec} can be intepreted as a metric recovery, which recover the underlying metric of the semantic manifold ~\autocite{hashimoto2016word}.

@@

When $w=1$ (Fig.~\ref{fig:supp:hyperparameter}), the flow can be written as
\begin{align}
	\label{eq:flow_w2v}
	\hat{T}_{ij} \propto P(i) P(j\given i) = \frac{P(i)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $P(j)$ is the fraction of $j$ in the data.
In general, calculating $Z_i$ is computationally expensive and there are two common approximations: hierarchical softmax \autocite{morin2005hierarchical} and negative sampling \autocite{mikolov2013word2vec}.
Due to its simplicity and performance, negative sampling is the most widely used strategy, which we also adopt in our study.

@@

Parameter $\gamma=1$ is a special choice that ensures that, when the embedding dimension is sufficiently large, there exists optimal in-vectors and out-vectors such that $\vect{v}_i = \vect{u}_i$ \autocite{levy2014neural}.
Setting $\gamma = 1$ and substituting $\vect{v}_i = \vect{u}_i$ into Eq.~\ref{eq:prob_w2v_ng0}, the flow predicted by {\it word2vec} is given by
\begin{align}
	\widehat{T}_{ij} \propto  N(i) P(j\given i) = \frac{NP(i)P(j)\exp(\vect{v}_j \cdot \vect{v}_{i})}{Z'_i}. \label{eq:flow_w2v_ng1}
\end{align}
where $N(i) = NP(i)$ is the frequency of location $i$ in the data, and $N=\sum_{i \in {\cal A}} N(i)$ is the sum of frequencies of all locations.

@@

In other words, \emph{word2vec with skip-gram negative sampling is equivalent to the gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model describes mobility flows from the given mass and locations, {\it word2vec}
finds locations that best explain the given mobility flow.
This makes \textit{word2vec} a powerful tool for learning models of mobility.

@@

In the remainder of the paper, we focus on scientific mobility and interrogate the geometric space generated by the neural embedding to shed light on the multi-faceted relationships between organizations.
To explore the topological structure of the embedding, we use a topology-based dimensionality reduction method (UMAP~\autocite{mcinnes2018umap}) to obtain a two-dimensional representation of the embedding space (Fig.~\ref{fig:projection}a).
By leveraging the unique characteristics of \textit{word2vec}, we are able to show the relationships between individual organizations, rather than aggregates such as nations or cities, this projection constitutes the largest and highest resolution ``map'' of scientific mobility to date.

>>>>

Here, we study the skip-gram negative sampling (SGNS), or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning models of language~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered}, networks~\autocite{grover2016node2vec, perozzi2014deepwalk, linzhuo2020hyperbolic} and trajectories\autocite{liu2016exploring, feng2017poi2vec, yao2018representing, cao2019habit2vec, crivellari2019motion, solomon2018predict}.
It works under the notion that a good representation should facilitate prediction, learning a mapping between words that can predict a target word based on its context (surrounding words).
The model is also computationally efficient, easy to use, robust to noise, and can encode relations between entities as geometric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis, hashimoto2016word}.
When applied to the trajectory data, each location is encoded into a single vector representation, and vectors relate to one another based on the likelihood of locations appearing adjacent to one another in the same trajectory. Also,  \textit{word2vec} can be intepreted as a metric recovery, which recover the underlying metric of the semantic manifold ~\autocite{hashimoto2016word}.

@@

When $w=1$ (Fig.~\ref{fig:supp:hyperparameter}), the flow can be written as
\begin{align}
	\label{eq:flow_w2v}
	\hat{T}_{ij} \propto P(i) P(j\given i) = \frac{P(i)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $P(j)$ is the fraction of location $j$'s frequency in the data.
In general, calculating $Z_i$ is computationally expensive and there are two common approximations: hierarchical softmax \autocite{morin2005hierarchical} and negative sampling \autocite{mikolov2013word2vec}.
Due to its simplicity and performance, negative sampling is the most widely used strategy, which we also adopt in our study.

@@

Parameter $\gamma=1$ is a special choice that ensures that, when the embedding dimension is sufficiently large, there exists optimal in-vectors and out-vectors such that $\vect{v}_i = \vect{u}_i$ \autocite{levy2014neural}.
Setting $\gamma = 1$ and substituting $\vect{v}_i = \vect{u}_i$ into Eq.~\ref{eq:prob_w2v_ng0}, the flow predicted by {\it word2vec} is given by
\begin{align}
	\widehat{T}_{ij} \propto  N(i) P(j\given i) = \frac{NP(i)P(j)\exp(\vect{v}_j \cdot \vect{v}_{i})}{Z'_i}. \label{eq:flow_w2v_ng1}
\end{align}
where $N$ is the sum of frequencies of all locations and $N(i) = NP(i)$ is the frequency of location $i$ in the data.

@@

In other words, \emph{word2vec with skip-gram negative sampling is mathematically equivalent to the gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model describes mobility flows from the given mass and locations, {\it word2vec}
estimates proper locations in the vector space which best explain the given mobility flow.
This makes \textit{word2vec} a powerful tool for learning models of mobility.

@@

In the remainder of the paper, we focus on scientific mobility and interrogate the geometric space generated by the neural embedding to shed light on the multi-faceted relationships between organizations.
To explore the topological structure of the embedding, we use a topology-based dimensionality reduction method (UMAP~\autocite{mcinnes2018umap}) to obtain a two-dimensional representation of the embedding space (Fig.~\ref{fig:projection}a).
By leveraging the unique characteristics of representation learning approach, we are able to show the relationships between individual organizations, rather than aggregates such as nations or cities, this projection constitutes the largest and highest resolution ``map'' of scientific mobility to date.
